---
title: "Figure 5: Pathway Analysis"
author: |
  | Dr Yue Zhao
  | Gilbert S. Omenn Department of Computational Medicine and Bioinformatics
  | University of Michigan
date: "6/16/2025"
output: 
  rmarkdown::html_document:
    theme: spacelab
    highlight: haddock
    code_folding: hide
    toc: true
    toc_float: true
    smooth_scroll: true
    number_sections: false
    self_contained: true  
---

```{r setup, include=FALSE}
library(here)
library(readxl)
library(dplyr)
library(reshape2)
library(ComplexHeatmap)
library(pheatmap)
library(RColorBrewer)
library(gdata)
library(magrittr)
library(circlize)
library(ggfortify)
library(pander)
library(ggsci)
library(ggpubr)
library(gplots)
library(gridExtra)
library(tidyverse)
library(gtsummary)
library(DT)
library(knitr)
library(kableExtra)
library(limma)
library(edgeR)
library(ggplot2)
library(rlang)
library(factoextra)
library(car)
library(pathview)
library(clusterProfiler)
library(org.Hs.eg.db)
library(msigdbr)
library(openxlsx)
library(ggh4x)
library(scales)
knitr::opts_knit$set(root.dir = "C:/Users/yuedz/OneDrive - Michigan Medicine/Documents/Project/ALS/Manuscript-RNA/Scripts_github/")
here::i_am("RNA_paper_Fig5.Rmd")
load("C:/Users/yuedz/OneDrive - Michigan Medicine/Documents/Project/ALS/RNA-seq/CaseVSControl/sample694/RNA_paper_pathway_yesCellPro/plot_no_yes_df3.rda")
```


# Fig.5a 

Enriched KEGG and Hallmark pathways in ALS cases versus controls shared after pathway analysis from primary DEGs (left) and DEGs adjusted (right) for cell proportions, shown overall (grey), in males (green), and females (yellow). Up- and downregulated pathways annotated at the bottom of the dot plot. 

```{r, fig.width=10, fig.height=6, warning=FALSE}
plot_common <- filter(plot_yes, Description%in%plot_no$Description)
plot_show <- filter(plot_yes, !(Description%in%plot_no$Description))
plot_disappear <- filter(plot_no, !(Description%in%plot_yes$Description))

plot_df3_common <- filter(plot_df3, Description%in%plot_common$Description)
plot_df3_show <- filter(plot_df3, Description%in%plot_show$Description)
plot_df3_disappear <- filter(plot_df3, Description%in%plot_disappear$Description)

# Calculate the frequency of each Description term for each level of comparison2
freq_table <- plot_df3_common %>% group_by(Description, comparison2) %>% summarise(freq = n(),.groups = "drop") %>% spread(key = comparison2, value = freq, fill = 0)

# Calculate the difference in frequencies
freq_table <- freq_table %>% mutate(diff_freq = `Adjust celltype` - `Not adjust celltype`)
plot_df3_common <- plot_df3_common %>% dplyr::left_join(freq_table %>% dplyr::select(Description, diff_freq), by = "Description")

# Order the data frame based on the difference in frequencies
plot_df3_common <- plot_df3_common %>%
  arrange(comparison2,comparison,desc(diff_freq)) %>%
  filter(!(category == "Hallmark" & Description == "Oxidative phosphorylation"))
plot_df3_common$Description<-factor(plot_df3_common$Description,levels=unique(plot_df3_common$Description))

# Figure 5A
ggplot(plot_df3_common, 
       aes(x=factor(method, levels = c("Up","Down")),
           y=Description)) + 
  geom_point(aes(size=Count, color=FDR)) +
  scale_colour_gradient(low="red", high="blue") +
  theme(axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle=45,hjust=0.9,size=12),
        axis.text.y = element_text(size=12),
        panel.background = element_rect(fill = "white",
                                        colour = "white"),
        panel.border = element_rect(color = "black", 
                                    fill = NA, 
                                    linewidth = 1),
        panel.grid.major = element_line(linewidth = 0.25, linetype = 'solid',
                                        colour = "lightgrey")) +
  scale_y_discrete(limits=rev) +
  facet_nested(. ~comparison2+comparison) +
  theme(strip.background=element_rect(color="grey30", fill="grey90")) +
  theme(strip.text.x = element_text(size = 12, colour = "black", angle = 0))
```



# Fig.5b 

Enriched KEGG and Hallmark pathways in ALS cases versus controls that appeared after pathway analysis from DEGs adjusted for cell proportions. 

```{r, fig.width=10, fig.height=5, warning=FALSE}
plot_df3_show2 <- plot_df3_show %>% filter(FDR<0.0001)
plot_df3_show3 <- plot_df3_show2
plot_df3_show3$comparison2 <- "Not adjust celltype"
plot_df3_show4 <- rbind.data.frame(plot_df3_show3, plot_df3_show2)
plot_df3_show4$comparison2<-factor(plot_df3_show4$comparison2,levels=c("Not adjust celltype","Adjust celltype"))

ggplot(plot_df3_show4, 
       aes(x=factor(method, levels = c("Up","Down")),
           y=Description)) + 
  geom_point(aes(size = ifelse(comparison2 == "Adjust celltype", Count, NA),
                 color = ifelse(comparison2 == "Adjust celltype", FDR, NA))) +
  scale_colour_gradient(name= "FDR", low="red", high="blue") +
  scale_size_continuous(name = "Count") +
  theme(axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle=45,hjust=0.9,size=12),
        axis.text.y = element_text(size=12),
        panel.background = element_rect(fill = "white",
                                        colour = "white"),
        panel.border = element_rect(color = "black", 
                                    fill = NA, 
                                    linewidth = 1),
        panel.grid.major = element_line(linewidth = 0.25, linetype = 'solid',
                                        colour = "lightgrey")) +
  scale_y_discrete(limits=rev, labels= label_wrap(42)) +
  facet_nested(. ~comparison2+comparison) +
  theme(strip.background=element_rect(color="grey30", fill="grey90")) +
  theme(strip.text.x = element_text(size = 12, colour = "black", angle = 0))

```

# Fig.5c

Enriched KEGG and Hallmark pathways in ALS cases versus controls that disappeared after pathway analysis from DEGs adjusted for cell proportions.

```{r, fig.width=8, fig.height=4,warning=FALSE}
plot_df3_disappear2<-plot_df3_disappear
plot_df3_disappear2$comparison2<-"Adjust celltype"
plot_df3_disappear3<-rbind.data.frame(plot_df3_disappear,plot_df3_disappear2)


ggplot(plot_df3_disappear3, 
       aes(x=factor(method, levels = c("Up","Down")),
           y=Description)) + 
  geom_point(aes(size = ifelse(comparison2 == "Not adjust celltype", Count, NA),
                 color = ifelse(comparison2 == "Not adjust celltype", FDR, NA))) +
  scale_colour_gradient(name= "FDR", low="red", high="blue") +
  scale_size_continuous(name = "Count") +
  theme(axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle=45,hjust=0.9,size=12),
        axis.text.y = element_text(size=12),
        panel.background = element_rect(fill = "white",
                                        colour = "white"),
        panel.border = element_rect(color = "black", 
                                    fill = NA, 
                                    linewidth = 1),
        panel.grid.major = element_line(linewidth = 0.25, linetype = 'solid',
                                        colour = "lightgrey")) +
  scale_y_discrete(limits=rev) +
  facet_nested(. ~comparison2+comparison) +
  theme(strip.background=element_rect(color="grey30", fill="grey90")) +
  theme(strip.text.x = element_text(size = 12, colour = "black", angle = 0))
```


